---
title: "Battlefield"
author:
  - name: Jean-Philippe Villemin
    affiliation:
      - Institut de Recherche en Cancérologie de Montpellier,
        Inserm, Montpellier, France 
    email: jean-philippe.villemin@inserm.fr
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: > 
  <p>`Battlefield` is a **Swiss-army toolkit designed to define
  and extract spatial spots from specific regions
  in spatial transcriptomics data.**
  </br></br>
  It provides low-level, modular utilities to delineate
  spatial regions of interest,  
  including **interfaces between clusters**, **intra-cluster layers**, and  
  **inter-cluster trajectories**.
  </br></br>
  These utilities are intended to be reused and composed within higher-level  
  analytical workflows and packages.
  </br></br>
  `Battlefield` supports sequencing-based spatial transcriptomics platforms such as  
  **10x Genomics Visium**, across multiple resolutions, including  
  Visium HD (binned).</br></br>
  Battlefield package version: `r packageVersion("Battlefield")`
output: 
  rmarkdown::html_vignette:
    self_contained: true
    toc: true
    toc_depth: 4
    highlight: pygments
    fig_height: 3
    fig_width: 3
    fig_caption: no
    code_folding: show
package: Battlefield
vignette: >
  %\VignetteIndexEntry{Battlefield-Main}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

```{r load-libs, message = FALSE,  warning = FALSE, results = FALSE}
library(Battlefield)
library(SpatialExperiment)
library(ggplot2)
library(dplyr)
library(tidyr)
library(pheatmap)
library(pals)
library(grid)
library(patchwork)
```
# Introduction  

`Battlefield` is a **Swiss-army toolkit designed to define
and extract spatial spots from specific regions
in spatial transcriptomics data.**
</br></br>
It provides low-level, modular utilities to delineate
spatial regions of interest,  
including **interfaces between clusters**, **intra-cluster layers**, and  
**inter-cluster trajectories**.
</br></br>
These utilities are intended to be reused and composed within higher-level  
analytical workflows and packages.
</br></br>
`Battlefield` supports sequencing-based spatial transcriptomics platforms such as  
**10x Genomics Visium**, across multiple resolutions, including  
Visium HD (binned).</br></br>

# What is it for?  

`Battlefield` provides four core functionalities to define and extract 
spatial transcriptomics spots from specific tissue regions, enabling 
the study of spatial organization under different biological contexts:

* **Interfaces between clusters** — identify and analyze boundary 
regions where distinct tissue or cell populations interact.

* **Intra-cluster layers** — characterize spatial layers
or gradients within a single cluster.

* **Inter-cluster trajectories**  — model spatial transitions 
across multiple clusters.

* **Cluster spatial neighborhood**  — report the cluster composition 
of a spatial neighborhood defined by k nearest spots, 
constrained by a distance threshold.


# Starting point

We start from a `SpatialExperiment` object that contains **(i)** 
spatial coordinates for each spot and **(ii)** a precomputed clustering
stored in colData().  

Battlefield operates on a simple spot-level table, 
so we first export a data.frame with four columns:  

* spot_id: spot/barcode identifier (colnames(spe))  
* x, y: spatial coordinates (spatialCoords(spe))  
* cluster: cluster label provided by the user (a column in colData(spe))  

Because the cluster annotation can have different names
depending on the workflow (e.g., cluster, 
seurat_clusters, BayesSpace, etc.), the user can specify
which colData() column should be used.

We  load a simulated Visium dataset (hexagonal grid) to illustrate 
the different functionalities that Battlefield provides.

```{r interfaces1, fig.dim = c(7, 3)}
# Load Visium data 
data("visium_simulated_spe")

df <- data.frame(
spot_id = colnames(visium_simulated_spe),
x = spatialCoords(visium_simulated_spe)[, 1],
y = spatialCoords(visium_simulated_spe)[, 2],
cluster = colData(visium_simulated_spe)$cluster
)

# Plot cluster distribution
ggplot(df, aes(x, y, fill = cluster)) +
geom_point(size = 3.2,colour = "grey", shape = 21) +
coord_equal() +
theme_minimal()+
labs(x = "", y = "") +
theme( axis.text = element_blank()) 
```

You can have acces to some basic information about the dataset
you are using (grid type,distance threshold(radius) recommonded, etc.).
```{r interfaces2, results='asis'}
# Detect grid type and get parameters
res <- detect_grid_type(df, verbose = TRUE)
params <- get_neighborhood_params(df, verbose = TRUE)
```

## Interfaces between clusters

Battlefield allows you to identify and extract **spatial interfaces** 
between clusters. 
Here we demonstrate how to:

1. Detect interfaces between two specific clusters (according to three modes 
  of selection called `inner`, `outer` or `both`)
2. Identify multi-interface spots all at once
3. Detect a same number of control spots relative to the cluster of interest

`Battlefield` offers several functionalities related to interfaces 
(aka borders) via the following fonctions metionned below:  

* ` select_border_spots`, between two clusters according to a specified mode
* ` build_all_borders`
* ` select_core_spots` , to select control spots within a cluster
   relative to its border
* ` build_all_cores`

You must have first defined borders before selecting core spots.

Here we show an example to detect interfaces between clusters 4 and 5.
We indentify border spots from cluster 4 towards cluster 5 that we
referenced as interface.  
We used the `outer` mode to select spots, meaning
that only spots from cluster 4 that are located on the outside of the
interface towards cluster 5 are selected.

```{r interfaces3, fig.dim = c(7, 3)}

# Select reverse : cluster 4 → 5
border_out <- select_border_spots(df, 
    cluster = 4, 
    interface = 5,
    mode = "outer")
knitr::kable(head(border_out))

# Structre data 
df_out <- df |>
mutate(is_border = spot_id %in% border_out$spot_id) |>
left_join(border_out |> 
select(spot_id, is_border_multiple),by = "spot_id") |>
mutate(is_border_multiple = coalesce(is_border_multiple, FALSE))


# Plot interface 4 → 5
ggplot(df_out, aes(x, y, color = cluster)) +
geom_point(size = 1) +
geom_point(data = subset(df_out, is_border & !is_border_multiple),
          color = "black", size = 2) +
geom_point(data = subset(df_out, is_border & is_border_multiple),
          color = "red", size = 2) +
coord_equal() +
theme_minimal()+
labs(x = "", y = "",color = "") +
theme( axis.text = element_blank()) 

```

We can also identify all interfaces at once
and visualize spots that are part of multiple interfaces.  Here we used
a `max_dist = 60` microns and `k = 6` neighbors to define the spatial neighborhood.
`is_border_multiple` indicates whether a spot is part of multiple interfaces and
is plotted with a different alpha in the example.

```{r interfaces4, fig.dim = c(7, 3)}
all_borders <- build_all_borders(df,max_dist = 60 , k = 6)
knitr::kable(head(all_borders))

# Structre data 
df2 <- df |>
left_join(all_borders |> 
select(spot_id, undirected_pair, is_border_multiple),by = "spot_id") |>
mutate(is_border = !is.na(undirected_pair),
        is_border_multiple = coalesce(is_border_multiple, FALSE))

# Plot all interfaces
ggplot(df2, aes(x, y)) +
geom_point(data = subset(df2, !is_border),
            color = "gray", size = 3.2) +
geom_point(data = subset(df2, is_border & is_border_multiple),
            aes(color = undirected_pair), size = 3.2, alpha = 0.4) +
geom_point(data = subset(df2, is_border & !is_border_multiple),
            aes(color = undirected_pair), size = 3.2) +
coord_equal() +
theme_minimal()+
labs(x = "", y = "",color = "") +
theme( axis.text = element_blank()) 
```
Here we plot the different
types of spots identified for the interface between clusters 3 and 4.
We identify border spots for both directions (3→4 and 4→3) 
and a relative selection of core spots within cluster 3 that can 
be used as control spots for further analysis.

```{r interfaces5, fig.dim = c(6, 5)}

# Get all pairs and detect grid
pairs_visium <- directed_cluster_interface_pairs(df$cluster)
params_visium <- get_neighborhood_params(df, verbose = FALSE)

# Define cluster pair of interest
cluster_A <- 3
cluster_B <- 4

# Select border spots for both directions
border_A_to_B <- subset(
build_all_borders(df, k = 6, pairs = pairs_visium),
cluster == cluster_A & interface == cluster_B
)

border_B_to_A <- subset(
build_all_borders(df, k = 6, pairs = pairs_visium),
cluster == cluster_B & interface == cluster_A
)

# Build all borders to identify inner spots
all_borders_visium <- build_all_borders(df, k = 6, pairs = pairs_visium)

# Select inner spots for cluster A
inner_A <- select_core_spots(df, all_borders_visium, 
                            cluster = cluster_A, 
                            interface = cluster_B, 
                            mode = "both")

# Create visualization dataframe with spot classification
df_final <- df |>
mutate(spot_type = "background") |>
mutate(spot_type = ifelse(spot_id %in% border_A_to_B$spot_id, 
                            "border_A_to_B", spot_type)) |>
mutate(spot_type = ifelse(spot_id %in% border_B_to_A$spot_id, 
                            "border_B_to_A", spot_type)) |>
mutate(spot_type = ifelse(spot_id %in% inner_A$spot_id, 
                            "inner_A", spot_type)) |>
as.data.frame()

# Plot: clusters with borders and core control spots
ggplot(df_final, aes(x, y, color = cluster)) +
geom_point(data = subset(df_final, spot_type == "background"), 
            size = 1.5, alpha = 0.6) +
geom_point(data = subset(df_final, spot_type == "border_A_to_B"), 
            color = "red", size = 3.5, shape = 4, stroke = 1.5) +
geom_point(data = subset(df_final, spot_type == "border_B_to_A"),
            color = "orange", size = 3.5, shape = 4, stroke = 1.5) +
geom_point(data = subset(df_final, spot_type == "inner_A"), 
            aes(color = NULL), 
            color = "darkgreen", size = 3.5, shape = 1, stroke = 1.5) +
coord_equal() +
theme_minimal()+
labs(x = "", y = "",color = "") +
theme( axis.text = element_blank()) 
```

## Inter-cluster trajectories  

 
```{r loading,eval=TRUE,fig.dim = c(7,3)}

expr <- as.numeric(assay(visium_simulated_spe, "counts")["FAKE_GENE", ])
test <- cbind(as.data.frame(colData(visium_simulated_spe)), df[, c("x", "y"), drop = FALSE])
test$expr <- expr


start_cluster <- "3"
end_cluster   <- "4"
top_n <- 8

centroids <- compute_centroids(df)

A <- centroids[centroids$cluster == start_cluster, c("x","y")]
B <- centroids[centroids$cluster == end_cluster,   c("x","y")]


(res <- build_one_trajectory(df, A, B, top_n = top_n, max_dist = NULL))

ggplot(test, aes(x, y)) +
geom_point(aes(color = expr),size = 1.6, alpha = 0.85) +
scale_color_gradient(
low = "blue",
high = "red"
) +
geom_path(
data = res,
aes(x = x, y = y),
color = "green", linewidth = 1.1
) +
geom_path(
data = res,
aes(x = x, y = y),
color = "black", linewidth = 1.1,
arrow = grid::arrow(type = "closed",
length = grid::unit(5, "mm"))
) +
geom_point(data = res$selected, aes(x, y), 
inherit.aes = FALSE, size = 2.2, color="green") +
coord_equal() +
theme_minimal()


(out <- build_similar_trajectories(df, A, B, top_n = top_n, n_extra = 2, side = "both"))

ggplot(test, aes(x, y)) +
geom_point(aes(color = expr),size = 1.6, alpha = 0.85) +
scale_color_gradient(
low = "blue",
high = "red"
) +
geom_path(data=out, aes(x=x, y=y, group=trajectory_id), 
inherit.aes = FALSE, color="green",linewidth=1,
arrow = grid::arrow(type = "closed", length = grid::unit(3, "mm"))) +
geom_point(data = out$lines, aes(x, y), 
inherit.aes = FALSE, size = 2.2, color="green") +
coord_equal() +
theme_minimal() 


meta <- out|>
transmute(
    spot_id = as.character(spot_id),
    trajectory = as.character(trajectory_id),
    progress = as.numeric(pos_on_seg)
) |>
group_by(trajectory) |>
arrange(progress, .by_group = TRUE) |>
mutate(
    index = seq_len(n())   # <- index 1..length ordered according to t
) |>
ungroup()

gene <- c("FAKE_GENE") 
expr <- assay(visium_simulated_spe, "counts")[gene, meta$spot_id]
meta$expr <- expr

mat <- meta |>
select(trajectory, index, expr) |>
tidyr::pivot_wider(names_from = index, values_from = expr) |>
as.data.frame()

rownames(mat) <- mat$trajectory
mat$trajectory <- NULL

pheatmap(
mat,
cluster_rows = FALSE,
cluster_cols = FALSE,
border_color = "white",
main = gene,angle_col=0,
col=pals::coolwarm(), scale="row",
cellwidth=25, cellheight=25 , na_col = "grey90"
)
```  



# Session Information

```{r session-info}
sessionInfo()
```