---
title: "Battlefield : </br> A Swiss-army to define and extract spatial spots
 for different tasks."
author:
  - name: Jean-Philippe Villemin
    affiliation:
    - Institut de Recherche en Cancérologie de Montpellier,
     Inserm, Montpellier, France 
    email: jean-philippe.villemin@inserm.fr
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
abstract: > 
  <p> Battlefield  is a Swiss-army toolkit originally developed 
  to define and extract spatial spots from specific tissue regions—such 
  as front regions, niche borders, invasive margins, 
  and cluster interfaces—using spatial transcriptomics data or clustered tissue maps.

    It has since been extended to support trajectory selection and layer inspection,
    and now provides a collection of low-level utilities for spatial transcriptomics analysis. 
    These utilities are primarily intended to be reused within higher-level analytical packages.

    Battlefield enables margin/trajectory/layer-aware spot selection, 
    allowing users to construct consistent comparison groups (e.g. front vs core or ordered spots along a trajectory).  

    It is designed to work with sequencing-based platforms
    such as Visium (standard and Visium HD).<br/><br/>
    Battlefield package version: `r packageVersion("Battlefield")`
output: 
  rmarkdown::html_vignette:
      self_contained: true
      toc: true
      toc_depth: 4
      highlight: pygments
      fig_height: 3
      fig_width: 3
      fig_caption: no
      code_folding: show
package: Battlefield
vignette: >
  %\VignetteIndexEntry{Battlefield-Main}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::knit_hooks$set(optipng = knitr::hook_optipng)
```

```{r load-libs, message = FALSE,  warning = FALSE, results = FALSE}
library(Battlefield)
library(SpatialExperiment)
library(ggplot2)
library(dplyr)
library(tidyr)
library(pheatmap)
library(pals)
library(grid)

```


# What is it for? 

`Battlefield` Battlefield is a Swiss-army toolkit originally developed 
to define and extract spatial spots
from specific tissue regions—such as front regions, niche borders, 
invasive margins and cluster interfaces—using spatial transcriptomics data 
or clustered tissue maps.
It has since been extended to support trajectory selection and layer inspection, 
and now provides a collection of low-level utilities 
for spatial transcriptomics analysis.
These utilities are primarily intended to be reused within
higher-level analytical packages.
It is designed to work with sequencing-based platforms
such as Visium at several resolutions and Visium HD(binned).

# Let's illustrate its main functionalities

There are a three mains functionalities in Battlefield to define and extract
spots from specific tissue regions.
- Margins
- Layers
- Trajectories


## Margins Example  
## Layers Example  
## Trajectories Example  

Here 
```{r loading,eval=TRUE,fig.dim = c(7,3)}
data("visium_simulated_spe")
head("visium_simulated_spe")
set.seed(12)

head(colData(visium_simulated_spe))

df <- data.frame(
  spot_id = colnames(visium_simulated_spe),
  x = spatialCoords(visium_simulated_spe)[, 1],
  y = spatialCoords(visium_simulated_spe)[, 2],
  cluster = colData(visium_simulated_spe)$cluster
)

expr <- as.numeric(assay(visium_simulated_spe, "counts")["FAKE_GENE", ])
test <- cbind(as.data.frame(colData(visium_simulated_spe)), df[, c("x", "y"), drop = FALSE])
test$expr <- expr


start_cluster <- "3"
end_cluster   <- "4"
top_n <- 8

centroids <- compute_centroids(df)

A <- centroids[centroids$cluster == start_cluster, c("x","y")]
B <- centroids[centroids$cluster == end_cluster,   c("x","y")]


(res <- spots_near_segment(df, A, B, top_n = top_n, max_dist = NULL))

ggplot(test, aes(x, y)) +
  geom_point(aes(color = expr),size = 1.6, alpha = 0.85) +
  scale_color_gradient(
  low = "blue",
  high = "red"
  ) +
  geom_path(
  data = res$selected,
  aes(x = x, y = y),
  color = "green", linewidth = 1.1,
  arrow = grid::arrow(type = "closed", length = grid::unit(3, "mm"))
  ) +
  geom_point(data = res$selected, aes(x, y), 
  inherit.aes = FALSE, size = 2.2, color="green") +
  coord_equal() +
  theme_minimal()


(out <- build_parallel_lines(df, A, B, top_n = top_n, n_extra = 2, side = "both"))

ggplot(test, aes(x, y)) +
   geom_point(aes(color = expr),size = 1.6, alpha = 0.85) +
   scale_color_gradient(
   low = "blue",
   high = "red"
   ) +
   geom_path(data=out$lines, aes(x=x, y=y, group=line_id), 
   inherit.aes = FALSE, color="green",linewidth=1,
   arrow = grid::arrow(type = "closed", length = grid::unit(3, "mm"))) +
   geom_point(data = out$lines, aes(x, y), 
   inherit.aes = FALSE, size = 2.2, color="green") +
   coord_equal() +
   theme_minimal() 


meta <- out$lines |>
  transmute(
    spot_id = as.character(spot_id),
    trajectory = as.character(line_id),
    progress = as.numeric(t)
  ) |>
  group_by(trajectory) |>
  arrange(progress, .by_group = TRUE) |>
  mutate(
    index = seq_len(n())   # <- index 1..length ordered according to t
  ) |>
  ungroup()

gene <- c("FAKE_GENE") 
expr <- assay(visium_simulated_spe, "counts")[gene, meta$spot_id]
meta$expr <- expr

mat <- meta |>
  select(trajectory, index, expr) |>
  tidyr::pivot_wider(names_from = index, values_from = expr) |>
  as.data.frame()

rownames(mat) <- mat$trajectory
mat$trajectory <- NULL

pheatmap(
  mat,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  border_color = "white",
  main = gene,angle_col=0,
  col=pals::coolwarm(), scale="row",
  cellwidth=10, cellheight=10 , na_col = "grey90"
)
```  


    
# Session Information

```{r session-info}
sessionInfo()
```